---
title: Crossing the Language Barrier
author: ''
date: '2020-12-10'
slug: blog-post-1
categories: []
tags: []
description: ''
---

# Python and R: Let's Close the Gap!
While both Python and R are really useful languages on their own, sometimes they work best together. In computational biology, it may be easier to visualize data through using "reticulate" so that Python and R can communicate with one another. It sounds challenging at first, but it's actually quite straightforward!

We'll start of with a simple example!
```{r}
#R Code Chuck 
# "reticulate" is activated in R
library(reticulate)

# the object seasons is defined
seasons <- "There are four seasons."
```

```{python}
#Python Code Chuck
seasons = "I think summer is my favorite!"

# r. allows Python to access R-defined objects
print(r.seasons,seasons) 
```

We can also do the opposite and access Python defined objects in R!

```{r}
# py. allows R to access Python-defined objects
cat(c(seasons,py$seasons))
```

# Let's Do Some Modeling!

The dataset "quakes" displays 1000 occurences of earthquakes near Fiji.

```{python}
import pandas as pd

quakes = pd.read_csv("https://vincentarelbundock.github.io/Rdatasets/csv/datasets/quakes.csv",index_col=0)

# displays first five rows
quakes.head()
```


Now let's model that data in R! The code below will generate a histogram displaying the frequency of different depths of earthquakes near Fiji!


```{r}
# generates histogram of the "depth" variable from the quakes dataset

hist(quakes$depth, xlab = "Depth (km)", main = "Distribution of Earthquake Depths near Fiji", right = F, col = " dark red")
```

Based on the histogram, we can observe a biomodal distribution!
